# R Basics

```{r}
#| include: false
library(dplyr)
library(dslabs)
data("gapminder")
```

## Core concepts

### Useful utility functions

A handful of functions you will reach for constantly when exploring any R session or dataset.

```{r}
#| eval: false

# R version info
version
major <- version$major
minor <- version$minor

# Package version
packageVersion('caret')

# Print column names in a reusable format
dput(colnames(data))
```

### Hardware benchmarking

The `benchmarkme` package lets you check your machine's performance relative to other R users — useful when deciding whether to run heavy computations locally or on a cluster.

```{r}
#| eval: false
library(benchmarkme)

# Standard benchmark (arithmetic, matrix operations, etc.)
res <- benchmark_std(runs = 3)
plot(res)

# RAM and CPU info
get_ram()
get_cpu()

# I/O benchmark (time to read a 50MB file)
res_io <- benchmark_io(runs = 1, size = 50)
plot(res_io)
```

### Core data formats

R has two main binary formats for storing objects:

- `.csv` — plain text, portable, human-readable
- `.rds` — R's native binary format; preserves object class exactly

```{r}
#| eval: false
read.csv("data.csv")     # returns a data.frame
readRDS("data.rds")      # returns whatever was saved
```

---

## Vectors

Vectors are the fundamental unit in R. Everything is built on them.

```{r}
#| eval: false
# Inspect any object
str(gapminder)
class(gapminder)
glimpse(gapminder)    # tidyverse equivalent

# Dimensions
dim(mpg)
head(mpg)
tail(mpg)

# Access a function from a specific package
dplyr::glimpse(gapminder)

# Combine elements
x <- c(1, 2, 3)

# Vector arithmetic is element-wise
c(1, 2, 3) * c(7, 6, 5)

# Sequences
seq(1, 5, by = 1)
1:5
```

### Functions applied to vectors

```{r}
#| eval: false
sort(x)       # sort ascending
rev(x)        # reverse order
unique(x)     # remove duplicates
table(x)      # frequency table

# Replicate
rep("ABC", times = 3)
```

### Indexing

R uses 1-based indexing. Be careful: `x[1]` returns the first element, not the second.

```{r}
#| eval: false
# Named vector
x <- c(e = 2.718, pi = 3.14, sqrt2 = 1.414)
x[c("e", "pi")]
x[c(1, 3)]

# Matrix indexing
mat <- matrix(1:6, nrow = 2)
mat[1, ]     # first row
mat[, 2]     # second column

# List access
my_list <- list(a = 1:3, b = "hello")
my_list$a
my_list[["a"]]
```

### Sorting and ordering

```{r}
#| eval: false
library(dslabs)
data("murders")

sort(murders$total)          # returns sorted values
order(murders$total)         # returns indices of sorted values
rank(murders$total)          # returns rank of each element

# Sort a data frame by a column
murders[order(murders$total), ]
```

### Type coercion

R will coerce types silently, which can cause subtle bugs. Always check your types.

```{r}
#| eval: false
b <- c(1, "b", 3)
class(b)        # "character" — 1 and 3 were coerced

# Explicit coercion
as.numeric(b)    # NAs introduced for non-numeric values
as.character(42)
as.factor(c("low", "medium", "high"))
as.data.frame(mat)
```

### Factors and ordered variables

Factors represent categorical variables. Ordered factors carry additional information about the direction of the levels.

```{r}
# Unordered factor (nominal)
survey <- c("M", "F", "F", "M", "M")
factor_survey <- factor(survey)
levels(factor_survey)

# Renaming levels — order must match the current order of levels
levels(factor_survey) <- c("Female", "Male")

# Ordered factor (ordinal)
temp <- c("Low", "Medium", "High", "Low", "High")
temp_fct <- factor(temp, levels = c("Low", "Medium", "High"), ordered = TRUE)
temp_fct
```

---

## Data frames

### Creation and basic inspection

```{r}
#| eval: false
df <- data.frame(
  name  = c("Alice", "Bob", "Carol"),
  age   = c(34, 28, 45),
  score = c(88.5, 72.0, 91.3)
)

head(df)
tail(df)
str(df)
summary(df)
dim(df)
names(df)         # column names
```

### Subsetting

```{r}
#| eval: false
# Column access — all equivalent
df$age
df[["age"]]
df[, "age"]

# Row subsetting
df[1, ]           # first row
df[2:3, ]         # rows 2 and 3
df[, 1:2]         # first two columns

# Logical subsetting
df[df$age > 30, ]
```

### Statistics

```{r}
#| eval: false
table(df$name)                         # frequency table
table(treatment_group, control_group)  # cross-tabulation
```

---

## File system utilities

```{r}
#| eval: false
getwd()                # current working directory
setwd("/path/to/dir")  # change working directory
list.files()           # files in current directory
ls()                   # objects in current environment

file.path("folder", "file.csv")   # portable path construction

data()                  # list built-in datasets
class(obj)              # type of object
args(mean)              # arguments of a function
installed.packages()    # list installed packages
packageVersion("shiny") # version of a specific package
```

---

## Writing functions

```{r}
get_top_names <- function(.year, .sex) {
  # Note: avoid naming arguments the same as column names to prevent
  # scoping conflicts inside dplyr pipelines.
  babynames::babynames |>
    dplyr::filter(year == .year, sex == .sex) |>
    dplyr::slice_max(prop, n = 10) |>
    dplyr::mutate(name = forcats::fct_inorder(name))
}
```

---

## Conditionals

```{r}
#| eval: false
# Vectorized if-else (tidyverse)
dplyr::if_else(condition, true_value, false_value)

# Base R
ifelse(x > 0, "positive", "non-positive")

# case_when for multiple conditions
dplyr::case_when(
  x > 10  ~ "high",
  x > 5   ~ "medium",
  .default = "low"
)
```
