# Data Import & I/O

```{r}
#| include: false
library(tidyverse)
```

This chapter covers reading and writing data in R, from plain-text files to Excel workbooks and SQL databases.

---

## Plain-text files

### The `readr` package (recommended)

`readr` functions are faster and more consistent than base R equivalents. They return tibbles and provide helpful parsing messages.

| Function | Equivalent to | Separator |
|---|---|---|
| `read_csv()` | `read.csv()` | Comma |
| `read_tsv()` | `read.delim()` | Tab |
| `read_csv2()` | `read.csv2()` | Semicolon (European) |
| `read_delim()` | `read.table()` | Any delimiter |
| `read_fwf()` | — | Fixed-width |
| `read_rds()` | `readRDS()` | R binary |

```{r}
#| eval: false
# Basic import
df <- read_csv("data.csv")

# Control parsing explicitly
df <- read_csv(
  "data.csv",
  col_types = list(
    age   = col_integer(),
    score = col_double(),
    group = col_factor()
  ),
  skip   = 1,    # skip the first row
  n_max  = 500   # read only 500 rows
)

# Read only specific columns
df <- read_csv("data.csv", col_types = cols_only(id = col_character(), age = col_integer()))
```

### Reading multiple files at once

```{r}
#| eval: false
# Stack multiple CSV files with a filename identifier column
files <- c("data/01-cohort.csv", "data/02-cohort.csv", "data/03-cohort.csv")
combined <- read_csv(files, id = "source_file")
```

### Manual data entry

```{r}
#| eval: false
# tibble() works column by column
tibble(
  id    = 1:3,
  group = c("A", "B", "A"),
  score = c(82.1, 74.5, 90.2)
)

# tribble() works row by row — easier to read for small tables
tribble(
  ~id, ~group, ~score,
    1,   "A",   82.1,
    2,   "B",   74.5,
    3,   "A",   90.2
)
```

### Column type cheatsheet

| Collector | Type |
|---|---|
| `col_double()` | Numeric (floating point) |
| `col_integer()` | Integer |
| `col_character()` | String |
| `col_factor()` | Categorical |
| `col_logical()` | TRUE / FALSE |
| `col_date()` | Date (ISO 8601) |
| `col_datetime()` | Date + time |
| `col_skip()` | Skip this column |

### Parsing helpers

```{r}
#| eval: false
# Extract a number from a messy string
parse_number("$1,234.56")   # → 1234.56
parse_number("12 patients") # → 12

# Fix a column containing "five" as text
df <- df |>
  mutate(age = parse_number(if_else(age == "five", "5", age)))
```

---

## Writing files

```{r}
#| eval: false
write_csv(df, "output/results.csv")
write_tsv(df, "output/results.tsv")
write_rds(df, "output/results.rds")        # preserves R object exactly

# Parquet — fast, compressed, interoperable with Python/Spark
arrow::write_parquet(df, "output/results.parquet")
```

---

## Excel files

### Reading with `readxl`

```{r}
#| eval: false
library(readxl)

# List available sheets
excel_sheets("workbook.xlsx")

# Read a specific sheet
df <- read_excel("workbook.xlsx", sheet = "Sheet1")

# Read all sheets at once
all_sheets <- lapply(
  excel_sheets("workbook.xlsx"),
  read_excel,
  path = "workbook.xlsx"
)

# Specify column names when the file has no header row
df <- read_excel(
  "data.xlsx",
  sheet     = 1,
  col_names = c("country", paste0("year_", 1960:1966))
)
```

### Writing with `openxlsx`

```{r}
#| eval: false
library(openxlsx)
write.xlsx(df, "output.xlsx", sheetName = "Results")
```

---

## SQL databases

R can connect to any SQL database via `DBI` with a database-specific driver.

```{r}
#| eval: false
library(DBI)
library(RMySQL)

# Open a connection
con <- dbConnect(
  MySQL(),
  dbname   = "clinical_db",
  host     = "server.hospital.fr",
  port     = 3306,
  user     = "analyst",
  password = Sys.getenv("DB_PASSWORD")  # never hard-code passwords
)

# List tables
dbListTables(con)

# Import a full table
patients <- dbReadTable(con, "patients")

# Import all tables at once
tables <- lapply(dbListTables(con), dbReadTable, conn = con)

# SQL query
results <- dbGetQuery(con, "
  SELECT name, admission_date
  FROM patients
  WHERE icu_los > 2
    AND year(admission_date) = 2022
")

# Chunked import (useful for large tables)
res <- dbSendQuery(con, "SELECT * FROM lab_results WHERE patient_id > 1000")
chunk <- dbFetch(res, n = 500)   # fetch 500 rows at a time
dbClearResult(res)

# Always disconnect when done
dbDisconnect(con)
```

---

## Input / output utilities

### Console and file output

```{r}
#| eval: false
# print() adds a [1] prefix
print(c(1, 2, 3))

# cat() is cleaner for formatted output; writes to file with the file= argument
cat("Hello, world\n")
cat(c(1, 2, 3), file = "output.txt", sep = "\n")
cat(c(4, 5, 6), file = "output.txt", sep = "\n", append = TRUE)

# sink() redirects all console output to a file
sink("log.txt")
summary(df)
sink()  # restore console output
```

### File system

```{r}
#| eval: false
list.files()                        # files in working directory
list.files(all.files = TRUE)        # include hidden files
list.files("data/", pattern = "*.csv", full.names = TRUE)
```

### Reading tabular text with base R

```{r}
#| eval: false
# read.table() is the general base R function; all other base readers wrap it
df <- read.table("data.txt", header = TRUE, sep = "\t", dec = ".")

# Shortcuts
read.csv("data.csv")              # sep = ",", dec = "."
read.csv2("data.csv")             # sep = ";", dec = ","
read.delim("data.txt")            # sep = "\t", dec = "."
```

---

## Extracting text from PDFs

```{r}
#| eval: false
library(pdftools)

url <- "https://example.com/paper.pdf"
temp <- tempfile(fileext = ".pdf")
download.file(url, temp)
text <- pdf_text(temp)
file.remove(temp)

# text is a character vector with one element per page
cat(text[1])
```
