# Data Wrangling with dplyr

```{r}
#| include: false
#| eval: true
library(tidyverse)
library(gapminder)
library(babynames)
data("gapminder")
```

dplyr is the core tool for data manipulation in R. Its verbs are organised around what they operate on: **rows**, **columns**, **groups**, or **tables**.

---

## The pipe

```{r}
#| eval: false
# Base R pipe (R ≥ 4.1)
data |> filter(year == 2007) |> summarise(mean_life = mean(lifeExp))

# Tidyverse pipe (magrittr) — equivalent
data %>% filter(year == 2007) %>% summarise(mean_life = mean(lifeExp))
```

---

## Row verbs

### `filter()` — keep rows matching a condition

```{r}
#| eval: false
# Single condition
gapminder |> filter(continent == "Africa")

# Multiple conditions (AND)
gapminder |> filter(continent == "Africa", year == 2012, lifeExp >= 70)

# OR
gapminder |> filter(continent == "Africa" | continent == "Asia")

# %in% for multiple values
years    <- c(1990, 2000, 2007)
countries <- c("Vietnam", "France")
gapminder |> filter(year %in% years, country %in% countries)
```

### `arrange()` — reorder rows

```{r}
#| eval: false
gapminder |> arrange(lifeExp)           # ascending
gapminder |> arrange(desc(lifeExp))     # descending
gapminder |> arrange(continent, desc(gdpPercap))  # multi-column
```

### `distinct()` — remove duplicate rows

```{r}
#| eval: false
# Unique rows overall
df |> distinct()

# Unique combinations of specific columns
df |> distinct(country, continent, .keep_all = TRUE)
```

### `slice_*()` — extract rows by position

```{r}
#| eval: false
df |> slice_max(lifeExp, n = 5)     # top 5 rows by life expectancy
df |> slice_min(gdpPercap, n = 3)   # bottom 3 rows by GDP
df |> slice_head(n = 10)             # first 10 rows
df |> slice_sample(n = 100)          # random 100 rows
```

---

## Column verbs

### `select()` — choose columns

```{r}
#| eval: false
# By name
df |> select(country, year, lifeExp)

# By position
df |> select(1:3)

# Helper functions
df |> select(starts_with("gdp"))
df |> select(ends_with("_mm"))
df |> select(contains("life"))
df |> select(matches("\\.t\\."))      # regex

# Everything except
df |> select(-country, -year)
df |> select(-(col1:col4))

# Type-based selection
df |> select(where(is.numeric))
df |> select(where(is.character))
```

### `mutate()` — add or transform columns

```{r}
#| eval: false
# New column at the end
gapminder |> mutate(gdp_total = gdpPercap * pop)

# Control placement
gapminder |> mutate(gdp_total = gdpPercap * pop, .after = year)
gapminder |> mutate(gdp_total = gdpPercap * pop, .before = 1)

# Keep only used columns
flights |>
  mutate(
    gain          = dep_delay - arr_delay,
    gain_per_hour = gain / (air_time / 60),
    .keep = "used"
  )

# case_when for conditional recoding
df |> mutate(
  age_group = case_when(
    age < 18  ~ "Paediatric",
    age < 65  ~ "Adult",
    .default  = "Elderly"
  )
)

# Rename a column via rename() inside a pipeline with a new name
games |> mutate(
  cat_label = case_when(
    category == "avg_us_viewers"     ~ "Average US viewers",
    category == "rating_household"   ~ "Household rating",
    category == "ad_cost"            ~ "Ad cost (USD)",
    .default                         = as.character(category)
  )
)
```

### `rename()` — rename columns

```{r}
#| eval: false
df |> rename(new_name = old_name)
df |> janitor::clean_names()   # snake_case all column names automatically
```

### `relocate()` — move columns

```{r}
#| eval: false
df |> relocate(score, .before = age)
df |> relocate(id, .after = last_col())
```

---

## Grouped operations

### `group_by()` + `summarise()`

```{r}
#| eval: true
gapminder |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarise(
    total_pop    = sum(pop),
    median_life  = median(lifeExp),
    mean_gdp     = mean(gdpPercap),
    n_countries  = n()
  )
```

```{r}
#| eval: false
# Multiple grouping variables
gapminder |>
  group_by(continent, year) |>
  summarise(
    median_gdp     = median(gdpPercap),
    median_lifeExp = median(lifeExp),
    .groups = "drop"
  )
```

### Useful summary functions

| Function | Description |
|---|---|
| `n()` | Count rows in group |
| `n_distinct(x)` | Count unique values of `x` |
| `mean(x, na.rm=TRUE)` | Mean |
| `median(x, na.rm=TRUE)` | Median |
| `sd(x, na.rm=TRUE)` | Standard deviation |
| `first(x)`, `last(x)`, `nth(x, 3)` | Positional values |

### `count()` — quick frequency tables

```{r}
#| eval: false
df |> count(group)                    # equivalent to group_by() + summarise(n = n())
df |> count(group, sort = TRUE)       # sorted descending
df |> count(group, wt = population)   # weighted count
```

### `top_n()` within groups

```{r}
#| eval: false
# Counties with highest total population in each metro category
counties |>
  group_by(state, metro) |>
  summarise(total_pop = sum(population)) |>
  top_n(1, total_pop) |>
  ungroup() |>
  count(metro)
```

### Gotcha: variable name conflicts in functions

When writing functions that use dplyr verbs, avoid naming function arguments the same as column names. Use a `.` prefix convention:

```{r}
#| eval: false
# BAD: filter(year == year) matches nothing
get_top_names_broken <- function(year, sex) {
  babynames |> filter(year == year, sex == sex)  # bug!
}

# GOOD: prefix function arguments with .
get_top_names <- function(.year, .sex) {
  babynames |> filter(year == .year, sex == .sex)
}
```

---

## Pivoting data

Tidy data has one variable per column, one observation per row, one value per cell. `pivot_longer()` and `pivot_wider()` move data between wide and long formats.

### `pivot_longer()` — wide → long

```{r}
#| eval: false
# Weekly chart positions stored in columns wk1, wk2, ..., wk76
billboard |>
  pivot_longer(
    cols       = starts_with("wk"),
    names_to   = "week",
    values_to  = "rank",
    values_drop_na = TRUE
  ) |>
  mutate(week = parse_number(week))
```

```{r}
#| eval: false
# Multiple variables encoded in column names
# e.g. columns like "dob_child1", "name_child1", "dob_child2", "name_child2"
household |>
  pivot_longer(
    cols      = !family,
    names_to  = c(".value", "child"),  # .value tells pivot to use the prefix as the new column name
    names_sep = "_",
    values_drop_na = TRUE
  )
```

### `pivot_wider()` — long → wide

```{r}
#| eval: false
cms_patient_experience |>
  pivot_wider(
    id_cols     = starts_with("org"),
    names_from  = measure_cd,
    values_from = prf_rate
  )
```

---

## Joins

```{r}
#| eval: false
# Left join: keep all rows from the left table
flights |>
  left_join(airlines, by = "carrier") |>
  rename(airline_name = name) |>
  left_join(airports, by = c("dest" = "faa")) |>
  rename(airport_name = name)

# Common join types
inner_join(x, y, by = "id")  # only matching rows in both
left_join(x, y, by = "id")   # all rows from x
right_join(x, y, by = "id")  # all rows from y
full_join(x, y, by = "id")   # all rows from both
anti_join(x, y, by = "id")   # rows in x not in y
```

---

## Data cleaning helpers

### Missing values

```{r}
#| eval: false
# Detect
is.na(df$age)
sum(is.na(df))              # total NAs
any(is.na(df))              # any NAs?
colSums(is.na(df))          # NAs per column

# Complete cases only
df[complete.cases(df), ]
na.omit(df)

# Replace specific values with NA
df$score[df$score == 999] <- NA
```

### Dates with `lubridate`

```{r}
#| eval: false
library(lubridate)

ymd("2024-03-15")
ymd_hms("2024-03-15 08:30:00")
dmy("15/03/2024")
```

### Strings with `stringr`

All `stringr` functions start with `str_` and take a string vector as first argument.

```{r}
#| eval: false
library(stringr)

str_trim("  excess whitespace  ")     # trim leading/trailing spaces
str_pad("123", width = 6, pad = "0") # "000123"
str_detect(x, "pattern")             # logical: does the pattern appear?
str_count(x, "a")                    # count occurrences
str_subset(x, "ICU")                 # keep strings matching pattern
str_replace(x, "old", "new")         # replace first match
str_replace_all(x, "old", "new")     # replace all matches
str_sub(x, 1, 4)                     # extract characters 1–4
str_length(x)                        # number of characters

# Splitting
str_split(x, fixed(" - "), n = 2, simplify = TRUE)
```

### String concatenation

```{r}
#| eval: false
# paste() default separator is a space; paste0() default is ""
paste("Hello", "world", sep = " ")
paste0("patient_", 1:3)           # "patient_1" "patient_2" "patient_3"

# str_c() from stringr: propagates NA (better for data pipelines)
str_c("Hello", NA)  # NA  (paste() would give "Hello NA")
str_c(c("a", "b", "c"), collapse = ", ")  # "a, b, c"
```

### Number formatting

```{r}
#| eval: false
format(3.14159, digits = 3)                # "3.14"
format(1234567, big.mark = ",")            # "1,234,567"
formatC(0.001234, format = "e", digits = 2) # "1.23e-03"
formatC(3.14159, format = "f", digits = 2)  # "3.14"
```

---

## Practical example: babynames

```{r}
#| eval: false
library(babynames)

# Trend for a single name
babynames |>
  filter(name == "Sophie") |>
  ggplot(aes(x = year, y = n, color = sex)) +
  geom_line()

# Year-normalised proportions and change over time
babynames |>
  group_by(year) |>
  mutate(year_total = sum(n)) |>
  ungroup() |>
  mutate(fraction = n / year_total) |>
  group_by(name) |>
  arrange(name, year) |>
  mutate(ratio = fraction / lag(fraction)) |>     # year-on-year growth
  filter(fraction >= 0.00001) |>
  slice_max(ratio, n = 1)
```

---

## Collapsing boolean columns into a single factor

A common pattern in clinical data: you have several mutually exclusive binary columns (e.g. admission reason) and want a single factor column instead.

```{r}
#| eval: false
tmp <- data.frame(
  respiratory = sample(c(TRUE, FALSE), 40, replace = TRUE),
  neurologic  = sample(c(TRUE, FALSE), 40, replace = TRUE),
  shock       = sample(c(TRUE, FALSE), 40, replace = TRUE)
)

tmp$reason <- apply(tmp[1:3], 1, function(row) names(tmp)[which(row)])
```

---

## Converting factors to numeric

```{r}
#| eval: false
# Convert all factor columns to numeric
is_factor <- sapply(df, is.factor)
df[is_factor] <- lapply(df[is_factor], function(x) as.numeric(as.character(x)))
```
