# Efficient R Programming

```{r}
#| include: false
library(microbenchmark)
```

Writing R that runs fast matters as soon as your datasets grow beyond a few thousand rows, or when you run simulations or bootstrap procedures. This chapter covers timing, memory, vectorisation, and parallelism.

---

## Timing code

### `system.time()`

Returns three numbers: `user` (CPU time for user code), `system` (CPU time for system calls), and `elapsed` (wall-clock time). For most purposes, `elapsed` is what you care about.

```{r}
colon_fn       <- function(n) 1:n
seq_default    <- function(n) seq(n)
seq_by         <- function(n) seq(1, n, by = 1)

n <- 1e8
system.time(colon_fn(n))
system.time(seq_default(n))
system.time(seq_by(n))
```

The colon operator is the fastest because it maps directly to a C primitive.

### `microbenchmark` — more precise comparisons

```{r}
library(microbenchmark)

n <- 1e6
microbenchmark(
  colon    = 1:n,
  seq_def  = seq(n),
  seq_by   = seq(1, n, by = 1),
  times    = 20
)
```

`microbenchmark` runs each expression many times and reports summary statistics (min, mean, median, max) in nanoseconds or microseconds.

---

## Memory and vector allocation

Growing a vector inside a loop is one of the most common performance mistakes in R. Each iteration copies the entire vector to a new, larger block of memory.

```{r}
n <- 10000

# BAD — grows on each iteration
growing <- function(n) {
  x <- NULL
  for (i in seq_len(n)) x <- c(x, rnorm(1))
  x
}

# GOOD — pre-allocate, then fill
pre_allocate <- function(n) {
  x <- numeric(n)
  for (i in seq_len(n)) x[i] <- rnorm(1)
  x
}

# BEST — use a vectorised function directly
vectorised <- function(n) rnorm(n)

microbenchmark(
  growing     = growing(n),
  pre_alloc   = pre_allocate(n),
  vectorised  = vectorised(n),
  times = 5
)
```

---

## Vectorisation

R's built-in functions are implemented in C or Fortran and execute on entire vectors at once. Whenever you find yourself writing a loop over a vector, ask whether a vectorised function already exists.

```{r}
n <- 1e6

# Non-vectorised: one call per element
x <- numeric(n)
for (i in seq_len(n)) x[i] <- rnorm(1)

# Vectorised: one call for all elements
x <- rnorm(n)
```

```{r}
microbenchmark(
  loop_version = {
    x <- numeric(n)
    for (i in seq_along(x)) x[i] <- rnorm(1)
  },
  vectorised = rnorm(n),
  times = 3
)
```

The speedup is typically 100–1000×.

### `rowSums()` and `&&` vs `&`

```{r}
#| eval: false
# rowSums() is faster than apply(mat, 1, sum)
rowSums(mat)
colSums(mat)
rowMeans(mat)

# && short-circuits (stops evaluating after the first FALSE)
# & always evaluates both sides
if (condition_1 && condition_2) { ... }   # use in if() statements
df |> filter(a & b)                       # use & for vector operations
```

---

## Matrices vs. data frames

Matrices hold a single data type; data frames can hold multiple. Subsetting a matrix is substantially faster.

```{r}
#| eval: false
n <- 1000
mat <- matrix(rnorm(n * 10), ncol = 10)
df  <- as.data.frame(mat)

microbenchmark(mat[, 1], df[, 1])
```

When all your data is numeric (e.g. a feature matrix for modelling), convert to a matrix before running computations.

---

## Profiling with `profvis`

`profvis` shows a flame graph of where time is spent in a block of code. Wrap any expression in `profvis({...})` to visualise it.

```{r}
#| eval: false
library(profvis)
library(ggplot2movies)

profvis({
  comedies <- movies[movies$Comedy == 1, ]
  plot(comedies$year, comedies$rating)
  model <- loess(rating ~ year, data = comedies)
  j <- order(comedies$year)
  lines(comedies$year[j], model$fitted[j], col = "red")
})
```

Look for wide bars (slow functions) and unexpected calls.

---

## Parallel computing

R is single-threaded by default. The `parallel` package lets you distribute independent tasks across multiple CPU cores.

### When to parallelise

Parallel computing helps when:

- Tasks are **independent** (no iteration $i$ depends on $i-1$) — Monte Carlo simulations, bootstrap resampling, cross-validation folds.
- Each task takes at least a few milliseconds (overhead makes it worthless for trivial tasks).

### Basic pattern with `parApply`

```{r}
#| eval: false
library(parallel)

# How many cores are available?
detectCores()

# Create a cluster (leave one core free for the OS)
cl <- makeCluster(detectCores() - 1)

# parApply is a parallel drop-in for apply()
result <- parApply(cl, matrix_data, 2, median)

stopCluster(cl)
```

### `parSapply` for list/vector inputs

```{r}
#| eval: false
library(parallel)

# A single bootstrap resample
bootstrap <- function(data) {
  idx <- sample(nrow(data), replace = TRUE)
  data[idx, ]
}

cl <- makeCluster(detectCores() - 1)

# Export custom functions and data to each worker
clusterExport(cl, c("bootstrap", "my_data"))

# Run 1000 bootstrap samples in parallel
results <- parSapply(cl, 1:1000, function(i) {
  b <- bootstrap(my_data)
  mean(b$outcome)
})

stopCluster(cl)
```

---

## The `apply` family (single-core)

The `apply` functions are cleaner alternatives to `for` loops when working with matrices or lists. They do not automatically speed things up — use them for readability.

| Function | Input | Output |
|---|---|---|
| `apply(X, 1, FUN)` | Matrix — apply FUN to each row | Vector or list |
| `apply(X, 2, FUN)` | Matrix — apply FUN to each column | Vector or list |
| `sapply(X, FUN)` | List/vector | Simplified (vector or matrix) |
| `lapply(X, FUN)` | List/vector | Always a list |
| `vapply(X, FUN, type)` | List/vector | Typed output (safer than `sapply`) |

```{r}
#| eval: false
mat <- matrix(1:12, nrow = 3)
apply(mat, 1, sum)   # row sums
apply(mat, 2, mean)  # column means

sapply(1:5, function(x) x^2)  # c(1, 4, 9, 16, 25)
```

---

## Error handling and reproducibility

### `reprex` — reproducible examples

When asking for help, always provide a minimal reproducible example. The `reprex` package formats it for GitHub or Stack Overflow automatically.

```{r}
#| eval: false
y <- 1:4
mean(y)
reprex::reprex()   # copies a formatted snippet to the clipboard
```

### `tryCatch` — handle errors gracefully

```{r}
#| eval: false
result <- tryCatch(
  expr  = { log(-1) },
  warning = function(w) {
    message("Warning caught: ", conditionMessage(w))
    NA_real_
  },
  error = function(e) {
    message("Error caught: ", conditionMessage(e))
    NA_real_
  }
)
```
