# Data Visualisation with ggplot2

```{r}
#| include: false
#| eval: true
library(ggplot2)
library(dplyr)
library(palmerpenguins)
library(gapminder)
library(ggthemes)
data("gapminder")
data("penguins")
```

ggplot2 uses a **layered grammar of graphics**: you build a plot by adding layers (geoms, scales, facets, themes) on top of a base `ggplot()` call. Extensions for specialised plot types are available at [exts.ggplot2.tidyverse.org](https://exts.ggplot2.tidyverse.org/gallery/).

---

## The basics

### Anatomy of a ggplot call

```{r}
#| eval: false
ggplot(
  data    = penguins,              # the data frame
  mapping = aes(x = flipper_length_mm, y = body_mass_g)  # global aesthetic mapping
) +
  geom_point(aes(color = species, shape = species)) +   # local aesthetic
  geom_smooth(method = "lm") +
  labs(
    title    = "Body mass vs. flipper length",
    subtitle = "Palmer Archipelago penguins",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    caption  = "Data: palmerpenguins package"
  ) +
  ggthemes::scale_color_colorblind()
```

### Global vs. local aesthetics

Aesthetics specified inside `ggplot(aes(...))` apply to **all** geoms. Aesthetics inside a specific `geom_*()` apply only to that layer.

```{r}
#| eval: true
#| message: false
# Global: both geom_point and geom_smooth share color = species
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

---

## Common geoms

### Scatter plot — `geom_point()`

```{r}
#| eval: true
gapminder |>
  filter(year == 2007) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  labs(x = "GDP per capita (log scale)", y = "Life expectancy")
```

### Line plot — `geom_line()`

```{r}
#| eval: true
#| message: false
gapminder |>
  group_by(year, continent) |>
  summarise(median_gdp = median(gdpPercap), .groups = "drop") |>
  ggplot(aes(x = year, y = median_gdp, color = continent)) +
  geom_line() +
  expand_limits(y = 0)
```

### Histogram — `geom_histogram()`

Takes a single `x` aesthetic. Use `binwidth` to control bin width or `bins` to set the number of bins.

```{r}
#| eval: true
gapminder |>
  filter(year == 1952) |>
  ggplot(aes(x = pop / 1e6)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  labs(x = "Population (millions, log scale)", y = "Count")
```

### Bar plot — `geom_col()` vs `geom_bar()`

- `geom_bar()` counts rows automatically (useful when data has one row per observation).
- `geom_col()` uses a `y` variable you supply (useful after summarising).

```{r}
#| eval: false
# geom_col() with summarised data
gapminder |>
  filter(year == 1952) |>
  group_by(continent) |>
  summarise(median_gdp = median(gdpPercap)) |>
  ggplot(aes(x = continent, y = median_gdp)) +
  geom_col()

# geom_bar() — reorder by frequency
penguins |>
  ggplot(aes(x = fct_infreq(species))) +
  geom_bar()
```

### Box plot — `geom_boxplot()`

```{r}
#| eval: false
penguins |>
  ggplot(aes(x = species, y = bill_depth_mm)) +
  geom_boxplot()
```

### Smooth lines — `geom_smooth()`

```{r}
#| eval: false
ggplot(data, aes(x = dose, y = response)) +
  geom_jitter(width = 0.05, height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"))
```

---

## Aesthetics reference

| Aesthetic | Controls |
|---|---|
| `color` / `colour` | Outline or line colour |
| `fill` | Interior fill colour |
| `size` | Point size or line width |
| `shape` | Point shape (0–25) |
| `alpha` | Transparency (0 = invisible, 1 = opaque) |
| `linetype` | Line type (solid, dashed, dotted …) |

---

## Scales

```{r}
#| eval: false
# Log-scale axes
scale_x_log10()
scale_y_log10()

# Manual colour scale
scale_color_manual(values = c("group1" = "#E41A1C", "group2" = "#377EB8"))

# Colorblind-safe palette (ggthemes)
ggthemes::scale_color_colorblind()

# Axis limits without removing data (use with care)
expand_limits(y = 0)

# Discrete x-axis labels
scale_x_discrete(labels = c("val1" = "Label 1", "val2" = "Label 2"))
```

---

## Faceting

`facet_wrap()` creates a grid of panels — one per level of a variable.

```{r}
#| eval: true
gapminder |>
  filter(year == 1952) |>
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) +
  geom_point(alpha = 0.7) +
  scale_x_log10() +
  facet_wrap(~continent)
```

---

## Labelling

```{r}
#| eval: false
labs(
  title    = "Main title",
  subtitle = "Subtitle",
  x        = "X-axis label",
  y        = "Y-axis label",
  color    = "Legend title",
  caption  = "Data source: ..."
)
```

---

## Overlaying multiple filtered geoms

When you want different subsets of the data on different layers, pass filtered data directly to each geom:

```{r}
#| eval: false
ggplot(data) +
  geom_line(
    data = data |> filter(group == "A"),
    aes(x = time, y = value), color = "steelblue"
  ) +
  geom_line(
    data = data |> filter(group == "B"),
    aes(x = time, y = value), color = "tomato"
  )
```

---

## Special characters in labels

Use Unicode escape sequences when you need special characters in axis labels or titles:

```{r}
#| eval: false
"m\u00B2"   # m²
"\u03B2"    # β (beta)
"\u2264"    # ≤
"\u2265"    # ≥
```

---

## Venn diagrams

### `ggVennDiagram`

```{r}
#| eval: false
library(ggVennDiagram)

set.seed(42)
gene_sets <- list(
  Study_A = sample(paste0("gene", 1:1000), 300),
  Study_B = sample(paste0("gene", 1:1000), 525),
  Study_C = sample(paste0("gene", 1:1000), 440)
)

ggVennDiagram(gene_sets) +
  scale_fill_gradient(low = "grey90", high = "tomato")
```

### `VennDiagram` (base-graphics approach)

```{r}
#| eval: false
library(VennDiagram)

# 3-set Venn from pre-counted values
draw.triple.venn(
  area1 = 500, area2 = 480, area3 = 420,
  n12 = 200, n13 = 180, n23 = 160, n123 = 80,
  category = c("PTSD", "Anxiety disorder", "Depression"),
  fill = c("#E41A1C", "#377EB8", "#4DAF4A"),
  alpha = 0.5
)
```

---

## Radar / spider charts

The `fmsb` package draws radar charts from a data frame where the first row is the maximum value per axis and the second row is the minimum.

```{r}
#| eval: false
library(fmsb)

# Build the data frame
scores <- as.data.frame(matrix(
  sample(2:20, 10, replace = TRUE), ncol = 10
))
colnames(scores) <- c("math", "biology", "music", "R", "stats",
                      "french", "physics", "english", "coding", "sport")

# Add max and min rows
scores <- rbind(rep(20, 10), rep(0, 10), scores)

# Default radar chart
radarchart(scores)

# Customised version
radarchart(
  scores,
  axistype  = 1,
  pcol      = rgb(0.2, 0.5, 0.5, 0.9),  # polygon colour
  pfcol     = rgb(0.2, 0.5, 0.5, 0.5),  # fill colour
  plwd      = 3,
  cglcol    = "grey", cglty = 1,
  axislabcol = "grey", caxislabels = seq(0, 20, 5),
  vlcex     = 0.8
)
```

---

## Clinical bar plots with custom labels

A pattern that comes up often in clinical papers: side-by-side or stacked bars with percentage labels.

```{r}
#| eval: false
# Example: LVEF distribution in a cardiac cohort
data_lvef <- data.frame(
  group      = c("LVEF < 30%", "LVEF 30–50%", "LVEF > 50%"),
  percentage = c(35, 42, 23)
)

ggplot(data_lvef, aes(x = group, y = percentage, fill = group)) +
  geom_col(width = 0.5, color = "black") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = "", y = "Percentage (%)") +
  theme_bw() +
  theme(legend.position = "none")
```
