# Descriptive Tables with gtsummary

```{r}
#| include: false
#| eval: true
library(gtsummary)
library(dplyr)
library(gapminder)
```

`gtsummary` produces publication-ready summary tables with a concise, pipe-friendly API. It handles continuous variables, categorical variables, grouped comparisons, and p-values automatically.

---

## The core function: `tbl_summary()`

`tbl_summary()` takes a data frame and returns a formatted summary table. By default it detects variable types and applies sensible statistics.

```{r}
#| eval: true
#| message: false
gapminder |>
  select(continent, pop, gdpPercap, lifeExp) |>
  tbl_summary(by = continent) |>
  add_n() |>
  add_p() |>
  bold_labels() |>
  modify_header(label = "**Variable**")
```

---

## Customising statistics

Use `statistic` to control what is shown for each variable type:

```{r}
#| eval: false
tbl_summary(
  data      = base,
  by        = PTSD_CLUSTER_STATUS,
  missing   = "no",
  statistic = list(
    all_continuous()  ~ "{mean} ({sd})",        # mean ± SD
    all_categorical() ~ "{n} / {N} ({p}%)"      # n / N (%)
  ),
  digits = all_continuous() ~ 2
)
```

Alternative statistics strings:

| String | Meaning |
|---|---|
| `"{median} ({p25}, {p75})"` | Median (IQR) |
| `"{mean} ({sd})"` | Mean ± SD |
| `"{n} ({p}%)"` | Count (%) |
| `"{min}, {max}"` | Range |

You can target specific variables by name:

```{r}
#| eval: false
statistic = list(
  age    ~ "{median} ({p25}, {p75})",
  weight ~ "{mean} ({sd})",
  group  ~ "{n} ({p}%)"
)
```

---

## Custom labels

```{r}
#| eval: false
tbl_summary(
  data  = base,
  label = list(
    AGE_PA         = "Patient's age (years)",
    IGSII          = "IGS-II score",
    SCORE_CHARLSON = "Charlson comorbidity index"
  )
)
```

---

## Adding information with `add_*()`

| Function | What it adds |
|---|---|
| `add_n()` | Column with total non-missing values |
| `add_p()` | P-values for group comparisons |
| `add_overall()` | An "Overall" column |
| `add_difference()` | Difference between groups with CI |
| `add_q()` | Q-values (adjusted p-values) |

```{r}
#| eval: false
base |>
  tbl_summary(by = group) |>
  add_n()        |>
  add_p()        |>
  add_overall()  |>
  add_difference()
```

---

## Formatting the table

```{r}
#| eval: false
tbl |>
  modify_header(label = "**Variable**") |>
  modify_caption("**Table 1. Patient characteristics**") |>
  bold_labels() |>
  italicize_levels()
```

---

## Full clinical Table 1 example

This is a complete, annotated Table 1 template as used in a clinical paper.

```{r}
#| eval: false
library(gtsummary)
library(dplyr)

table1 <- base |>
  # Recode factor scores as numeric where needed
  mutate(
    across(c(QUALITE_COM_EQUIPE, AISE_EXPRIM_RESSENTI, IMPLICATION_DEC_MED,
             ENTOURAGE, EXCLUSION_SOCIALE, STRESS_FIN), as.numeric)
  ) |>
  tbl_summary(
    by        = PTSD_CLUSTER_STATUS,
    missing   = "no",
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = all_continuous() ~ 2,
    label  = list(
      AGE_PA         = "Patient's age (years)",
      IGSII          = "IGS-II severity score",
      SCORE_CHARLSON = "Charlson comorbidity index"
    )
  ) |>
  add_n()          |>   # non-missing N per variable
  add_p()          |>   # between-group p-value
  add_overall()    |>   # total column
  add_difference() |>   # difference with 95% CI
  modify_header(label = "**Variable**") |>
  modify_caption("**Table 1: Patient characteristics by PTSD cluster status**") |>
  bold_labels()

table1
```

---

## Regression tables with `tbl_regression()`

`gtsummary` can also format model output into publication-ready regression tables.

```{r}
#| eval: false
library(gtsummary)

model <- glm(ICU_mortality ~ age + SOFA + mv + vasopressors,
             data = icu_data, family = "binomial")

tbl_regression(
  model,
  exponentiate = TRUE,   # report OR instead of log-odds
  label = list(
    age          = "Age (years)",
    SOFA         = "SOFA score",
    mv           = "Mechanical ventilation",
    vasopressors = "Vasopressors"
  )
) |>
  add_global_p() |>
  bold_p() |>
  bold_labels()
```
