# Regression & Generalised Linear Models

```{r}
#| include: false
library(tidyverse)
library(broom)
```

Generalised linear models (GLMs) extend linear regression to outcomes that are not normally distributed. They share a common framework: a **distribution family**, a **link function**, and a **linear predictor**.

---

## Linear models

```{r}
#| eval: false
model <- lm(lifeExp ~ gdpPercap + year, data = gapminder)
summary(model)
coef(model)
confint(model)
```

A linear model is a special case of a GLM with a Gaussian family and identity link.

---

## Generalised linear models

```{r}
#| eval: false
# General syntax
glm(formula = y ~ x1 + x2, data = df, family = "binomial")

# Families
glm(y ~ x, data, family = "gaussian")    # same as lm()
glm(y ~ x, data, family = "binomial")    # logistic regression
glm(y ~ x, data, family = "poisson")     # Poisson regression

# Remove the intercept (estimate a coefficient for each group)
glm(y ~ group - 1, data, family = "binomial")
```

---

## Poisson regression

Poisson regression is appropriate for count outcomes, especially when counts are small and contain many zeros.

- Coefficients are on the **log scale** → exponentiate to get the rate ratio.
- Poisson coefficients are **multiplicative**, not additive.

```{r}
#| eval: false
pois_model <- glm(n_events ~ dose + group, data = df, family = "poisson")

# Exponentiate to get rate ratios
exp(coef(pois_model))

# Tidy output with confidence intervals
broom::tidy(pois_model, exponentiate = TRUE, conf.int = TRUE)
```

### Plotting a Poisson fit

```{r}
#| eval: false
ggplot(df, aes(x = dose, y = cell_count)) +
  geom_jitter(width = 0.05, height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"))
```

---

## Logistic regression

Logistic regression models a binary outcome $Y \in \{0, 1\}$.

**Model:**  
$\text{logit}(p) = \log\left(\frac{p}{1-p}\right) = \beta_0 + \beta_1 x$

**Inverse logit** (predicted probability):  
$p = \frac{1}{1 + e^{-x}}$

### Fitting

Three equivalent data formats:

```{r}
#| eval: false
# 1. Long format (one row per individual, 0/1 outcome)
lr1 <- glm(died ~ age + sofa, data = patients_long, family = "binomial")

# 2. Wide format (successes and failures as separate columns)
lr2 <- glm(cbind(deaths, survivors) ~ age, data = patients_wide, family = "binomial")

# 3. Proportion with weights
lr3 <- glm(mortality_rate ~ age, data = patients_agg, family = "binomial",
           weights = n_total)
```

### Link functions

The default is **logit**. The **probit** link is an alternative based on the normal CDF.

```{r}
#| eval: false
glm(y ~ x, family = binomial(link = "logit"))    # default
glm(y ~ x, family = binomial(link = "probit"))
```

Logit is preferred for outlier-robustness (heavier tails); probit assumes normality.

### Interpreting odds ratios

```{r}
#| eval: false
# Extract and exponentiate coefficients
coefs <- coef(model)
exp(coefs)               # odds ratios

# Using broom::tidy — cleaner
broom::tidy(model, exponentiate = TRUE, conf.int = TRUE)
```

| OR | Interpretation |
|---|---|
| > 1 | Exposure increases the odds of the outcome |
| = 1 | No effect |
| < 1 | Exposure decreases the odds of the outcome |

### Visualising logistic regression

```{r}
#| eval: false
ggplot(df, aes(x = age, y = icu_death)) +
  geom_jitter(width = 0, height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(y = "Probability of ICU death")
```

---

## Multiple regression

Rule of thumb: at least **10 events per predictor** to avoid overfitting.

```{r}
#| eval: false
model <- glm(
  died ~ age + sofa + mv + vasopressors + creatinine,
  data   = icu_data,
  family = "binomial"
)

summary(model)
```

### Model formula syntax

| Formula | Effect |
|---|---|
| `y ~ x` | Intercept + slope for x |
| `y ~ x - 1` | Slope for x, no intercept |
| `y ~ x1 + x2` | Multiple predictors |
| `y ~ x1 * x2` | Main effects + interaction |
| `y ~ x1:x2` | Interaction only |
| `y ~ .` | All columns as predictors |

### Factor reference levels

The **first level** of a factor is used as the reference by default (alphabetical order). Change it with `relevel()`:

```{r}
#| eval: false
df$status <- relevel(factor(df$status), ref = "Stable")
```

### Design matrix

```{r}
#| eval: false
model.matrix(~ size)             # reference intercept + contrasts
model.matrix(~ size - 1)         # intercept for each level
model.matrix(~ size + color - 1) # multiple factors, each with own intercept
```

---

## Extracting model results

```{r}
#| eval: false
coef(model)                     # coefficients
confint(model)                  # confidence intervals
predict(model, newdata = df)    # predictions (on link scale)
predict(model, newdata = df, type = "response")  # predictions (probability scale)

# Tidy output (broom package)
broom::tidy(model)
broom::glance(model)            # goodness-of-fit stats
broom::augment(model)           # original data + fitted values + residuals
```

---

## Model assumptions

For logistic regression the key assumptions are:

- **Linearity in the logit**: the log-odds is a linear function of continuous predictors.
- **Independence of observations**: no clustering or repeated measures (without special handling).
- **No severe multicollinearity**: correlated predictors inflate standard errors.
- **Adequate sample size**: ≥ 10 events per predictor.

Watch for **overdispersion** (more variance than the model expects) and **Simpson's paradox** (an important confounding variable missing from the model).

---

## Simulating binary data

```{r}
#| eval: false
# Bernoulli: n = number of trials, size = 1 (single coin flip each)
rbinom(n = 100, size = 1, prob = 0.3)

# Binomial: n = 1, size = 100 (one draw of 100 coin flips)
rbinom(n = 1, size = 100, prob = 0.3)

# Convert logit-scale predictions to probability
p <- plogis(0.5)   # = exp(0.5) / (1 + exp(0.5))
rbinom(n = 10, size = 1, prob = p)
```

---

## Correlations

```{r}
#| eval: false
cor(df$var1, df$var2)
cor(df, method = "spearman")        # rank correlation
cor.test(df$var1, df$var2)          # with p-value and CI

# Visualise correlation matrix
library(ggcorrplot)
df |>
  select(where(is.numeric)) |>
  cor(use = "pairwise.complete.obs") |>
  ggcorrplot(
    hc.order   = TRUE,
    type       = "upper",
    lab        = TRUE,
    method     = "circle",
    colors     = c("tomato2", "white", "springgreen3")
  )
```
