# Survival Analysis

```{r}
#| include: false
library(survival)
library(survminer)
library(ggplot2)
```

Survival analysis handles time-to-event outcomes. The key challenge is **censoring**: some patients have not yet experienced the event at the time of analysis. Standard regression methods cannot handle this correctly — survival methods can.

---

## Core concepts

- **Event**: the outcome of interest (death, ICU discharge, relapse, etc.)
- **Censored observation**: patient has not experienced the event by the end of follow-up (or was lost to follow-up)
- **Survival function** $S(t)$: probability of surviving beyond time $t$
- **Hazard function** $h(t)$: instantaneous rate of the event at time $t$, given survival to $t$

---

## The `Surv` object

Every survival analysis starts by creating a `Surv` object that packages the time and event status together.

```{r}
#| eval: false
# time: follow-up time (numeric)
# event: 1 = event occurred, 0 = censored
sobj <- Surv(time = patients$time, event = patients$status)

# Inspect: + indicates censored observations
sobj[1:10]
summary(sobj)
```

---

## Kaplan–Meier estimator

The Kaplan–Meier estimator is the non-parametric gold standard for estimating the survival function from observed data.

```{r}
#| eval: false
# Overall KM estimate
km <- survfit(Surv(time, event) ~ 1, data = patients)

# By group
km_by_group <- survfit(Surv(time, event) ~ treatment, data = patients)

# Inspect the survfit object
str(km)
km$surv        # estimated survival probabilities
km$time        # times at which the estimate changes
km$n.risk      # number at risk at each time
```

### Plotting with `ggsurvplot`

```{r}
#| eval: false
ggsurvplot(
  km,
  data             = patients,
  palette          = c("#E41A1C", "#377EB8"),
  surv.median.line = "hv",       # horizontal + vertical lines at median
  conf.int         = TRUE,
  risk.table       = TRUE,       # add risk table below
  cumevents        = TRUE,
  tables.height    = 0.25,
  legend.labs      = c("Control", "Treatment"),
  xlab             = "Time (days)",
  ylab             = "Survival probability"
)
```

### Comparing multiple KM curves

```{r}
#| eval: false
km_a <- survfit(Surv(time, event) ~ 1, data = cohort_a)
km_b <- survfit(Surv(time, event) ~ 1, data = cohort_b)

ggsurvplot_combine(
  list(CohortA = km_a, CohortB = km_b),
  data = patients
)
```

### Log-rank test

```{r}
#| eval: false
# Test whether survival differs between groups
survdiff(Surv(time, event) ~ treatment, data = patients)
```

---

## Cox proportional hazards model

The Cox model estimates how covariates affect the hazard rate, without specifying the baseline hazard function.

$h(t|X) = h_0(t) \cdot e^{\beta_1 X_1 + \beta_2 X_2 + \ldots}$

Coefficients are **log hazard ratios**. Exponentiate them to get hazard ratios (HR).

```{r}
#| eval: false
cox <- coxph(Surv(time, event) ~ age + sofa + mv + vasopressors,
             data = icu_data)

summary(cox)

# Extract tidy results
broom::tidy(cox, exponentiate = TRUE, conf.int = TRUE)
```

### Interpreting hazard ratios

| HR | Interpretation |
|---|---|
| > 1 | Higher hazard (shorter time to event) |
| = 1 | No association |
| < 1 | Lower hazard (longer time to event) |

### Checking the proportional hazards assumption

The Cox model assumes that the hazard ratio between groups is **constant over time**.

```{r}
#| eval: false
ph_test <- cox.zph(cox)
print(ph_test)       # p < 0.05 signals a violation
plot(ph_test)        # Schoenfeld residual plots
```

---

## Weibull parametric model

The Weibull model specifies the baseline hazard explicitly, which allows predictions beyond the observed follow-up time.

```{r}
#| eval: false
library(survival)

weibull_model <- survreg(Surv(time, event) ~ age + group,
                         data = patients, dist = "weibull")
summary(weibull_model)
```

---

## Propensity score methods

In observational ICU studies, treatment is rarely randomised. Propensity scores can balance covariates between groups.

```{r}
#| eval: false
library(MatchIt)

# Estimate propensity score and perform 1:1 matching
match_obj <- matchit(
  treatment ~ age + sofa + mv + vasopressors + creatinine,
  data   = icu_data,
  method = "nearest",
  ratio  = 1
)

# Extract matched data
matched_data <- match.data(match_obj)

# Fit Cox model in matched cohort
cox_matched <- coxph(Surv(time, event) ~ treatment,
                     data = matched_data, weights = weights)
```

---

## Useful packages

| Package | Purpose |
|---|---|
| `survival` | Core survival analysis (Surv, survfit, coxph, survreg) |
| `survminer` | ggplot2-based survival plots |
| `MatchIt` | Propensity score matching |
| `timeROC` | Time-dependent ROC curves |
| `cmprsk` | Competing risks (Fine–Gray model) |

---

## Quick reference: building blocks

```{r}
#| eval: false
# Create Surv object
sobj <- Surv(time, event)

# KM estimate
km <- survfit(sobj ~ group, data = df)
print(km)
summary(km, times = c(30, 90, 180, 365))   # survival at specific time points

# Log-rank test
survdiff(sobj ~ group, data = df)

# Cox model
cox <- coxph(sobj ~ age + group + sofa, data = df)
exp(coef(cox))      # hazard ratios
exp(confint(cox))   # confidence intervals

# KM plot
ggsurvplot(km, data = df, risk.table = TRUE, conf.int = TRUE)
```
