# Contingency Tables

```{r}
#| include: false
library(tidyverse)
```

Contingency tables summarise the joint distribution of two or more categorical variables. They are used to test independence and to estimate measures of association.

---

## Creating a contingency table

### From individual data

```{r}
#| eval: false
table(df$exposure, df$outcome)
```

### From pre-counted data

```{r}
# Direct matrix input
j0 <- c(6, 39)
j7 <- c(27, 18)

mat <- matrix(cbind(j0, j7), ncol = 2, byrow = FALSE)
colnames(mat) <- c("Day 0", "Day 7")
rownames(mat) <- c("Ratio < threshold", "Ratio ≥ threshold")

as.table(mat)
```

---

## Chi-square test

```{r}
#| eval: false
chisq.test(mat)
```

For small cell counts (expected < 5), use Fisher's exact test instead:

```{r}
#| eval: false
fisher.test(mat)
```

---

## McNemar's test for paired data

McNemar's test is the appropriate test when the two groups are **paired** (e.g. before/after measurements in the same patient, or matched pairs).

```{r}
# Sparse 2×2 table
data_sparse <- matrix(c(4, 2, 23, 16), ncol = 2, byrow = TRUE)
colnames(data_sparse) <- c("J7_below", "J7_above")
rownames(data_sparse) <- c("J0_below", "J0_above")

mcnemar.test(data_sparse, correct = FALSE)
```

### Odds ratio via `epibasix`

```{r}
#| eval: false
library(epibasix)

# Gives OR, confidence interval, and p-value
mcNemar(data_sparse, alpha = 0.05, digits = 3)
```

---

## Margin totals

```{r}
#| eval: false
margin.table(mat, 1)   # row totals
margin.table(mat, 2)   # column totals
prop.table(mat)        # proportions
prop.table(mat, 1)     # row proportions
prop.table(mat, 2)     # column proportions
```
