# Introduction to Bayesian Data Analysis

Bayesian inference updates a **prior distribution** (what we believe before seeing data) with observed data to produce a **posterior distribution** (our updated beliefs).

$$P(\theta | \text{data}) \propto P(\text{data} | \theta) \cdot P(\theta)$$

---

## Core vocabulary

| Term | Definition |
|---|---|
| Prior | $P(\theta)$ — distribution representing beliefs before data |
| Likelihood | $P(\text{data} \| \theta)$ — probability of the data given a parameter value |
| Posterior | $P(\theta \| \text{data})$ — updated beliefs after seeing data |
| Credible interval | Interval that contains the true parameter value with a given probability |

---

## Probability distributions in Bayesian analysis

### Simulating from a binomial process

```{r}
# 100 Bernoulli trials with p = 0.42
n_trials    <- 100
prop_success <- 0.42

# Method 1: using runif
data <- as.numeric(runif(n_trials) < prop_success)
sum(data)

# Method 2: using rbinom
rbinom(n = 1,   size = 100, prob = 0.42)   # total successes in 100 trials
rbinom(n = 100, size = 1,   prob = 0.42)   # 100 individual Bernoulli draws
```

### The Beta distribution

The Beta distribution is the natural prior for a probability parameter $p \in [0, 1]$.

- **Beta(1, 1)** = uniform prior (all values of $p$ equally likely)
- Increasing $\alpha$ pulls the distribution toward 1; increasing $\beta$ pulls it toward 0
- Larger $\alpha + \beta$ → narrower distribution (more prior certainty)

```{r}
#| eval: false
# Uninformative prior
beta_uniform <- rbeta(1000, shape1 = 1, shape2 = 1)
hist(beta_uniform, main = "Beta(1,1) - Uniform prior", xlab = "p")

# Informative prior: believe click-through rate ~5%
beta_informative <- rbeta(1000, shape1 = 5, shape2 = 95)
hist(beta_informative, main = "Beta(5, 95) - Informative prior", xlab = "p")
```

---

## A complete Bayesian workflow

This example models the click-through rate of an advertisement.

```{r}
#| eval: false
set.seed(42)
n_draws       <- 100000
n_ads_shown   <- 100
observed_clicks <- 13

# Step 1: Sample from the prior
proportion_clicks <- rbeta(n_draws, shape1 = 5, shape2 = 95)

# Step 2: Simulate outcomes under the prior
n_visitors <- rbinom(n_draws, size = n_ads_shown, prob = proportion_clicks)

prior <- data.frame(proportion_clicks, n_visitors)

# Step 3: Condition on observed data (posterior ≈ those simulations matching observation)
posterior <- prior[prior$n_visitors == observed_clicks, ]

# Step 4: Summarise the posterior
cat("Posterior median:", median(posterior$proportion_clicks), "\n")
cat("95% credible interval:", quantile(posterior$proportion_clicks, c(0.025, 0.975)), "\n")
cat("P(rate > 0.10):", mean(posterior$proportion_clicks > 0.10), "\n")
```

```{r}
#| eval: false
# Visualise prior vs posterior
par(mfrow = c(2, 1))
hist(prior$proportion_clicks,     xlim = c(0, 0.25), main = "Prior", xlab = "p")
hist(posterior$proportion_clicks, xlim = c(0, 0.25), main = "Posterior", xlab = "p")
par(mfrow = c(1, 1))
```

---

## Interpreting the posterior

```{r}
#| eval: false
median(posterior)               # point estimate
quantile(posterior, c(0.05, 0.95))   # 90% credible interval
sum(posterior > 0.07) / length(posterior)  # P(p > 0.07)
```

### Credible interval vs. confidence interval

- A **95% credible interval** says: "There is a 95% probability that the true parameter lies in this interval." This is the intuitive frequentist interpretation that people mistakenly apply to confidence intervals.
- A **95% confidence interval** says: "If we repeated this experiment many times, 95% of the intervals constructed this way would contain the true value."

---

## Bayesian vs. frequentist decision-making

| Question | Frequentist | Bayesian |
|---|---|---|
| "Is the effect real?" | p < 0.05 threshold | P(effect > 0) from posterior |
| Uncertainty | Confidence interval | Credible interval |
| Prior information | Not used | Formally incorporated |
| Small samples | Problematic | Priors can stabilise |

---

## Recommended packages

| Package | Purpose |
|---|---|
| `brms` | Bayesian regression using Stan |
| `rstanarm` | Pre-compiled Stan models, drop-in replacement for `lm`/`glm` |
| `bayesplot` | Visualisation of Bayesian model outputs |
| `tidybayes` | Tidy tools for extracting posterior draws |
