# Factor Analysis & Structural Equation Modelling

```{r}
#| include: false
library(psych)
library(lavaan)
```

Factor analysis is a set of techniques for discovering latent constructs (factors) that explain correlations among observed variables (items). It underpins the validation and scoring of psychometric scales, including instruments commonly used in clinical research such as the HADS.

---

## Key concepts

| Concept | Definition |
|---|---|
| Latent variable | An unobserved construct inferred from observed items (e.g. "Anxiety") |
| Factor loading | Correlation between an item and a factor |
| Communality | Proportion of an item's variance explained by the factors |
| EFA | **Exploratory** FA — no prior theory about which items load on which factors |
| CFA | **Confirmatory** FA — tests a specific hypothesised factor structure |

---

## Exploratory Factor Analysis (EFA)

### When to use EFA

Use EFA when you do not have a strong prior theory about the factor structure, or when you are developing a new scale.

### Running an EFA

```{r}
#| eval: false
library(psych)

# Single-factor EFA
efa_1 <- fa(hads_data)
efa_1
efa_1$loadings

# Multi-factor EFA
efa_3 <- fa(hads_data, nfactors = 3)
efa_3$loadings
head(efa_3$scores)

# Path diagram
fa.diagram(efa_3, labels = colnames(hads_data))
```

### Descriptive statistics for items

```{r}
#| eval: false
describe(hads_data)      # mean, SD, skew, kurtosis for each item
error.bars(hads_data)    # error bar plot
error.dots(hads_data)    # dot plot with error bars
```

### Determining the number of factors

**Eigenvalue criterion** (Kaiser rule): retain factors with eigenvalue > 1.

```{r}
#| eval: false
# Scree plot
item_cor   <- cor(hads_data, use = "pairwise.complete.obs")
eigen_vals <- eigen(item_cor)
eigen_vals$values

psych::scree(item_cor, factors = FALSE)
```

---

## Reliability

### Cronbach's alpha

Measures internal consistency — the extent to which items in a scale measure the same construct. A commonly cited threshold is α > 0.80 for good reliability.

```{r}
#| eval: false
alpha_result <- psych::alpha(hads_data)
alpha_result$total$raw_alpha

# Split-half reliability
psych::splitHalf(hads_data)
```

### Correlation matrix

```{r}
#| eval: false
# Base R
lowerCor(hads_data)
corr.test(hads_data, use = "pairwise.complete.obs", method = "pearson", adjust = "none")

# Correlogram with ggcorrplot
library(ggcorrplot)
p_mat <- cor_pmat(hads_data, use = "pairwise.complete.obs")

hads_data |>
  cor(use = "pairwise.complete.obs") |>
  ggcorrplot(
    hc.order    = TRUE,
    type        = "upper",
    lab         = TRUE,
    p.mat       = p_mat,
    sig.level   = 0.01,
    insig       = "blank",
    colors      = c("tomato2", "white", "springgreen3")
  )
```

---

## Splitting data for EFA/CFA

Best practice: run EFA on half the data, CFA on the other half.

```{r}
#| eval: false
set.seed(42)
N       <- nrow(scale_data)
idx_efa <- sample(seq_len(N), floor(0.5 * N))
idx_cfa <- setdiff(seq_len(N), idx_efa)

data_efa <- scale_data[idx_efa, ]
data_cfa <- scale_data[idx_cfa, ]
```

---

## Confirmatory Factor Analysis (CFA)

CFA tests a theoretically specified factor structure using the `lavaan` package.

### Model specification

```{r}
#| eval: false
library(lavaan)

# Two-factor model (lavaan syntax)
model_hads <- "
  Depression =~ HADS_DEP_Q8 + HADS_DEP_Q9 + HADS_DEP_Q10 +
                HADS_DEP_Q11 + HADS_DEP_Q12 + HADS_DEP_Q13 + HADS_DEP_Q14
  Anxiety    =~ HADS_ANX_Q1 + HADS_ANX_Q2 + HADS_ANX_Q3 +
                HADS_ANX_Q4 + HADS_ANX_Q5 + HADS_ANX_Q6 + HADS_ANX_Q7
"

fit <- cfa(model_hads, data = hads_data, std.lv = TRUE)
summary(fit, fit.measures = TRUE, standardized = TRUE)
```

### Model fit indices

| Index | Good fit |
|---|---|
| CFI (Comparative Fit Index) | > 0.90 |
| TLI (Tucker–Lewis Index) | > 0.90 |
| RMSEA | < 0.05 |
| SRMR | < 0.08 |

```{r}
#| eval: false
fitMeasures(fit, c("cfi", "tli", "rmsea", "srmr", "aic", "bic"))
```

### Comparing two models

```{r}
#| eval: false
# Likelihood-ratio test (only valid for nested models on the same data)
anova(fit_short, fit_full)

# BIC comparison (lower = better)
fitMeasures(fit_short, "bic")
fitMeasures(fit_full, "bic")
```

### McDonald's omega

Omega is a more accurate reliability estimate than Cronbach's alpha when factor loadings are unequal.

```{r}
#| eval: false
omega_result <- psych::omega(hads_data, nfactors = 2)
omega_result$omega.tot
```

---

## Cross-validation of a CFA model

```{r}
#| eval: false
library(cvTools)

k    <- 5
set.seed(123)
folds <- cvFolds(nrow(data_cfa), K = k)

cv_results <- data.frame(Fold = 1:k, CFI = NA, RMSEA = NA, SRMR = NA)

for (i in seq_len(k)) {
  test_idx  <- which(folds$which == i)
  train_dat <- data_cfa[-test_idx, ]
  fit_cv    <- cfa(model_hads, data = train_dat, std.lv = TRUE)
  fm        <- fitMeasures(fit_cv, c("cfi", "rmsea", "srmr"))
  cv_results[i, c("CFI", "RMSEA", "SRMR")] <- fm
}

colMeans(cv_results[, -1], na.rm = TRUE)
```

---

## Finding optimal scale cutoffs

When shortening a scale, you need to derive new cutoffs for the reduced score. The approach below finds the cutoff on the short scale that maximises Cohen's kappa agreement with the full-scale categories.

```{r}
#| eval: false
library(irr)

# Assume full-scale scores and categories are already computed
# Find optimal binary cutoff for the shortened scale
kappa_vals <- sapply(0:12, function(thresh) {
  short_case <- short_scores >= thresh
  if (length(unique(short_case)) == 2) {
    kappa2(data.frame(full_case, short_case))$value
  } else NA
})

best_cutoff <- which.max(kappa_vals) - 1   # adjust for 0-indexing
cat("Best cutoff:", best_cutoff, "| Kappa:", max(kappa_vals, na.rm = TRUE))
```

### ROC-based cutoff selection

```{r}
#| eval: false
library(pROC)

roc_obj   <- roc(full_case, short_scores, direction = "<")
auc(roc_obj)

best_coords <- coords(roc_obj, x = "best", best.method = "youden",
                      ret = c("threshold", "sensitivity", "specificity"))
print(best_coords)
```
