# Machine Learning in R

```{r}
#| include: false
library(tidyverse)
```

This chapter covers practical machine learning workflows in R, focusing on the tools most relevant for clinical and research applications.

---

## The `caret` package

`caret` (Classification And REgression Training) provides a unified interface for training and evaluating hundreds of models. It handles cross-validation, hyperparameter tuning, and pre-processing consistently across methods.

```{r}
#| eval: false
library(caret)

# Split into training and test sets
set.seed(42)
train_idx <- createDataPartition(df$outcome, p = 0.8, list = FALSE)
train_dat <- df[ train_idx, ]
test_dat  <- df[-train_idx, ]

# Define cross-validation scheme
ctrl <- trainControl(
  method    = "cv",
  number    = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

# Train a model (e.g. random forest)
model <- train(
  outcome ~ .,
  data      = train_dat,
  method    = "rf",
  trControl = ctrl,
  metric    = "ROC"
)

print(model)
varImp(model)   # variable importance

# Predict on test set
preds <- predict(model, newdata = test_dat)
confusionMatrix(preds, test_dat$outcome)
```

---

## Gradient boosting with XGBoost

XGBoost is the go-to algorithm for structured tabular data. It builds an ensemble of decision trees sequentially, each correcting the errors of the previous one.

### Cross-validation to find the optimal number of trees

```{r}
#| eval: false
library(xgboost)

cv <- xgb.cv(
  data              = as.matrix(train_features),
  label             = train_labels,
  nrounds           = 200,
  nfold             = 5,
  objective         = "reg:squarederror",  # or "binary:logistic"
  eta               = 0.1,     # learning rate
  max_depth         = 6,
  early_stopping_rounds = 10,
  verbose           = 0
)

# Find optimal number of trees
elog     <- cv$evaluation_log
ntrees   <- elog$iter[which.min(elog$test_rmse_mean)]
cat("Optimal trees:", ntrees)
```

### Training and prediction

```{r}
#| eval: false
model_xgb <- xgboost(
  data      = as.matrix(train_features),
  label     = train_labels,
  nrounds   = ntrees,
  objective = "reg:squarederror",
  eta       = 0.1,
  max_depth = 6,
  verbose   = 0
)

# Predict
test_data$pred <- predict(model_xgb, as.matrix(test_features))

# Evaluate (regression: RMSE)
test_data |>
  mutate(residuals = outcome - pred) |>
  summarise(rmse = sqrt(mean(residuals^2)))
```

### Visualise predictions vs. actuals

```{r}
#| eval: false
ggplot(test_data, aes(x = pred, y = outcome)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, color = "tomato") +
  labs(x = "Predicted", y = "Observed")
```

---

## Model evaluation

### Classification metrics

```{r}
#| eval: false
library(pROC)

roc_obj <- roc(true_labels, predicted_probs)
auc(roc_obj)
plot(roc_obj)

# Confusion matrix (caret)
confusionMatrix(predicted_classes, true_classes)
```

### Regression metrics

```{r}
#| eval: false
# RMSE
sqrt(mean((y_pred - y_true)^2))

# MAE
mean(abs(y_pred - y_true))

# R²
1 - sum((y_true - y_pred)^2) / sum((y_true - mean(y_true))^2)
```

---

## One-hot encoding

Many ML algorithms require numeric inputs. Factor variables must be converted to binary dummy columns.

```{r}
#| eval: false
library(caret)

# Create dummy variables
dummies    <- dummyVars(~ ., data = df)
df_encoded <- predict(dummies, newdata = df) |> as.data.frame()
```

Manual implementation:

```{r}
#| eval: false
# For a single factor column
model.matrix(~ factor_col - 1, data = df)
```

---

## K-means clustering

```{r}
#| eval: false
set.seed(42)
km_model <- kmeans(df_numeric, centers = 3, nstart = 25)

# Cluster assignments
km_model$cluster

# Cluster centres
km_model$centers

# Add cluster labels back to data
df$cluster <- factor(km_model$cluster)
```

### Choosing $k$

```{r}
#| eval: false
# Elbow method: plot within-cluster sum of squares vs. k
wcss <- sapply(1:10, function(k) kmeans(df_numeric, k, nstart = 25)$tot.withinss)
plot(1:10, wcss, type = "b", xlab = "k", ylab = "Within-cluster SS")
```

---

## Useful packages summary

| Package | Purpose |
|---|---|
| `caret` | Unified ML interface, cross-validation |
| `xgboost` | Gradient boosted trees |
| `randomForest` | Random forests |
| `glmnet` | Regularised regression (LASSO, ridge, elastic net) |
| `pROC` | ROC curves and AUC |
| `tidymodels` | Modern tidyverse-compatible ML framework |

---

## The `tidymodels` framework

`tidymodels` is the tidyverse successor to `caret`. It separates preprocessing (recipes), model specification, and fitting into distinct, composable steps.

```{r}
#| eval: false
library(tidymodels)

# 1. Preprocessing recipe
rec <- recipe(outcome ~ ., data = train_dat) |>
  step_impute_median(all_numeric_predictors()) |>
  step_normalize(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

# 2. Model specification
lr_spec <- logistic_reg(penalty = tune(), mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

# 3. Workflow
wf <- workflow() |>
  add_recipe(rec) |>
  add_model(lr_spec)

# 4. Tune with cross-validation
cv_folds <- vfold_cv(train_dat, v = 10)
tune_res  <- tune_grid(wf, resamples = cv_folds, grid = 20)

# 5. Finalise and fit
best_params <- select_best(tune_res, metric = "roc_auc")
final_fit   <- finalize_workflow(wf, best_params) |> fit(train_dat)

# 6. Evaluate on test set
final_fit |>
  augment(test_dat) |>
  roc_auc(outcome, .pred_Yes)
```
