# Building Interactive Apps with Shiny

```{r}
#| include: false
library(shiny)
library(ggplot2)
```

Shiny lets you build interactive web applications directly from R. No HTML, CSS, or JavaScript knowledge is required to get started — though you can add them for advanced customisation.

---

## Anatomy of a Shiny app

Every Shiny app has three components:

1. **UI** (`ui`) — defines the layout and appearance (frontend)
2. **Server** (`server`) — contains the computational logic (backend)
3. **`shinyApp()`** — combines them into a runnable app

```{r}
#| eval: false
library(shiny)

ui <- fluidPage()  # minimal empty UI

server <- function(input, output, session) {}  # empty server

shinyApp(ui = ui, server = server)
```

Run the app with `runApp("path/to/app/")` or by clicking **Run App** in RStudio.

---

## The workflow: input → server → output

```
User interacts with input widget
    ↓
input$id changes value
    ↓
server recomputes output$id
    ↓
Output renders in the UI
```

- Inputs are **always defined in the UI** and accessed in the server as `input$inputId`.
- Outputs are **rendered in the server** and displayed in the UI via `*Output()` functions.

---

## UI elements

### Layout functions

| Function | Purpose |
|---|---|
| `fluidPage()` | Fluid layout (adjusts to browser width) |
| `fixedPage()` | Fixed-width layout |
| `navbarPage()` | Page with navigation tabs |
| `sidebarLayout()` | Two-column layout with sidebar + main area |
| `sidebarPanel()` | Sidebar column |
| `mainPanel()` | Main column |
| `tabsetPanel()` + `tabPanel()` | Tabbed panels within a page |

### Standard layout with sidebar

```{r}
#| eval: false
ui <- fluidPage(
  titlePanel("My Clinical Dashboard"),
  sidebarLayout(
    sidebarPanel(
      textInput("name", "Patient name:"),
      sliderInput("age", "Age:", min = 0, max = 120, value = 50),
      selectInput("group", "Group:", choices = c("Control", "Treated"))
    ),
    mainPanel(
      plotOutput("survival_plot"),
      tableOutput("summary_table")
    )
  )
)
```

### Input widgets

| Widget | Function | Value type |
|---|---|---|
| Text field | `textInput()` | `character` |
| Numeric field | `numericInput()` | `numeric` |
| Slider | `sliderInput()` | `numeric` |
| Dropdown | `selectInput()` | `character` |
| Radio buttons | `radioButtons()` | `character` |
| Checkbox | `checkboxInput()` | `logical` |
| Multiple checkboxes | `checkboxGroupInput()` | `character vector` |
| Date picker | `dateInput()` | `Date` |
| Date range | `dateRangeInput()` | `Date[2]` (accessed via `[1]` and `[2]`) |
| Action button | `actionButton()` | integer (click count) |

```{r}
#| eval: false
# Slider with animation button
sliderInput("year", "Year:", min = 2000, max = 2024, value = 2010, animate = TRUE)

# Dropdown with grouped choices
selectInput(
  "species",
  "Select species:",
  choices = list(
    Penguins = list("Adelie", "Chinstrap", "Gentoo"),
    Others   = list("Emperor", "King")
  )
)

# Text input with placeholder
textInput("query", label = "", placeholder = "Search patient ID...")
```

### Output elements

| Display | UI function | Server render function |
|---|---|---|
| Plot | `plotOutput()` | `renderPlot({ ... })` |
| Table | `tableOutput()` | `renderTable({ ... })` |
| Text | `textOutput()` | `renderText({ ... })` |
| Verbatim text | `verbatimTextOutput()` | `renderPrint({ ... })` |
| Data frame | `dataTableOutput()` | `renderDataTable({ ... })` |
| UI element | `uiOutput()` | `renderUI({ ... })` |

---

## Server logic

```{r}
#| eval: false
server <- function(input, output, session) {

  # Render a plot that depends on the input
  output$survival_plot <- renderPlot({
    df_filtered <- patients |>
      filter(group == input$group, age <= input$age)

    ggplot(df_filtered, aes(x = time, y = survival)) +
      geom_line() +
      labs(title = paste("Group:", input$group))
  })

  # Render a summary table
  output$summary_table <- renderTable({
    patients |>
      filter(group == input$group) |>
      summarise(n = n(), median_los = median(los))
  })

}
```

---

## Reactive programming

Shiny uses **reactive expressions** to avoid unnecessary recomputation. A reactive expression caches its value and only re-runs when its inputs change.

```{r}
#| eval: false
server <- function(input, output, session) {

  # Reactive expression: filtered data (recomputes only when input$group changes)
  filtered_data <- reactive({
    patients |> filter(group == input$group)
  })

  # Multiple outputs can share the same reactive expression
  output$plot  <- renderPlot({ ggplot(filtered_data(), ...) })
  output$table <- renderTable({ filtered_data() })

}
```

### `observeEvent()` — respond to button clicks

```{r}
#| eval: false
observeEvent(input$run_analysis, {
  # This block runs only when the button is clicked
  output$result <- renderText(paste("Analysis run with n =", input$n))
})
```

---

## Running examples

The `shiny` package comes with built-in example apps:

```{r}
#| eval: false
runExample("01_hello")
runExample("02_text")       # tables and data frames
runExample("03_reactivity") # reactive expressions
runExample("05_sliders")    # slider widgets
runExample("06_tabsets")    # tabbed panels
runExample("09_upload")     # file upload
runExample("10_download")   # file download
```

---

## Complete working example

```{r}
#| eval: false
library(shiny)
library(babynames)
library(ggplot2)

ui <- fluidPage(
  titlePanel("Baby Name Explorer"),
  sidebarLayout(
    sidebarPanel(
      textInput("name", "Enter a name:", value = "Emma")
    ),
    mainPanel(
      plotOutput("trend", height = "400px")
    )
  )
)

server <- function(input, output, session) {
  output$trend <- renderPlot({
    babynames |>
      filter(name == input$name) |>
      ggplot(aes(x = year, y = prop, color = sex)) +
      geom_line(linewidth = 1) +
      labs(
        title = paste("Popularity of", input$name, "over time"),
        x = "Year", y = "Proportion of births"
      ) +
      theme_minimal()
  })
}

shinyApp(ui = ui, server = server)
```

---

## Deployment

| Platform | Command / method |
|---|---|
| Local (RStudio) | **Run App** button or `runApp()` |
| shinyapps.io (cloud) | `rsconnect::deployApp()` |
| Posit Connect | Deploy via IDE or `rsconnect` |
| Docker | Use `rocker/shiny` base image |

---

## Useful packages

| Package | Purpose |
|---|---|
| `shinydashboard` | Dashboard layouts with sidebar, header, boxes |
| `bslib` | Modern Bootstrap themes and UI components |
| `DT` | Interactive data tables (`renderDT`, `DTOutput`) |
| `plotly` | Interactive ggplot-to-plotly conversion |
| `shinyWidgets` | Additional input widgets |
| `reactable` | Fast, interactive tables |
